Spring haul-out behavior of seals in the Bering and Chukchi seas

### 2024-05-31 analysis/paper/London_HauloutBehavior.Rmd

knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  include = FALSE,
  fig.path = "../figures/",
  dpi = 600,
  fig.width = 6,
  fig.asp = 0.618,
  out.width = "100%",
  tab.topcaption = TRUE
)

>>>>

knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  include = FALSE,
  fig.path = "../figures/",
  dpi = 500,
  fig.width = 6,
  fig.asp = 0.618,
  out.width = "100%",
  tab.topcaption = TRUE
)

### 2024-05-31 analysis/paper/London_HauloutBehavior.tex

% Pandoc citation processing
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

@@

\hypertarget{introduction}{%
\section*{Introduction}\label{introduction}}
\addcontentsline{toc}{section}{Introduction}

@@

\hypertarget{methods}{%
\section*{Methods}\label{methods}}
\addcontentsline{toc}{section}{Methods}

\hypertarget{data-collection}{%
\subsection*{Data collection}\label{data-collection}}
\addcontentsline{toc}{subsection}{Data collection}

For this study we used haul-out behavior data and location estimates from
bio-loggers deployed on bearded, ribbon, and spotted seals in the Bering,
Chukchi, and western Beaufort seas by multiple organizations as part of
collaborative investigations from 2005 through 2021. Seals were captured using
nets and bio-loggers were attached during studies based in coastal communities
or on research ships. Ship-based capture events occurred during spring near the
southern ice edge in the Bering Sea between 2005 and 2018. Land-based capture
events occurred between 2005 and 2020 from May to October, generally between the
Alaska coastal communities of Scammon Bay in the Bering Sea, Utqiaġvik
in the Chukchi Sea, and Nuiqsut in the Beaufort Sea (Supplemental
Material, \protect\hyperlink{s1}{S1}). Data from additional
deployments along the Kamchatka peninsula in the western Bering Sea are also
included. We refer readers to Figure \ref{fig:deployplot} and the primary literature
for detailed capture and bio-logger attachment methods (see also Supplemental
Material, \protect\hyperlink{s1}{S1}).

\begin{figure}
\includegraphics[width=1\linewidth]{../figures/deployplot-1} \caption{\textbf{Initial bio-logger deployment areas for bearded, ribbon, and spotted seals in the study between 2005 and 2020 across multiple regions within the Bering, Chukchi, and western Beaufort seas.} \linebreak Solid regions shown for each species are minimum concave polygons buffered by 60 km for enhanced visibility. The larger, shaded region indicates the spatial extent of aerial survey effort to date in the Bering, Chukchi, and Beaufort seas. Deployments were initiated across a range of months but only haul-out data from 1 March to 15 July were included in the analysis. See Figure \ref{fig:dataMap} for the spatial distribution of observed data used in the study and Supplemental Materials \ref{tab:deploymentTable} for additional deployment details.}\label{fig:deployplot}
\end{figure}

We created a subset of haul-out behavior data from
250 bio-loggers deployed on
35 bearded,
110 ribbon, and
105 spotted seals
to include only records from 1 March to 15 July between
2005 and
2021. Bio-loggers were of the
`SPLASH' or `SPOT' family of tags developed by Wildlife Computers (Redmond,
Washington, USA). Deployments consisted of either a single `SPLASH' device, a
single `SPOT' device, or both types. Devices were either adhered to the hair on
the seal or attached through the rear flipper inter-digital webbing. The use of
bio-loggers adhered to the back or head provides some benefits over flipper
mounted devices (e.g.~increased satellite transmittal rates, locations at sea)
but these fall off during the following annual molt, which, depending on
deployment date, limits the duration of haul-out data they provide especially
during the focus months of our study. Additionally, bio-loggers attached to the
head or dorsal region are often dry while the seal is floating at the surface,
inducing a slight positive bias in the hourly percent-dry values reported by the
bio-logger. For this study, in cases where both bio-logger types were deployed,
we preferred hourly percent-dry observations from the flipper tag. All data were
transmitted by the deployed instruments via the Argos satellite network and
location data were either derived from Argos transmissions or transmitted
FastLoc GPS data.

Sex and age class (non-dependent \emph{young-of-the-year}, sexually immature
\emph{subadults}, and mature \emph{adults}) were estimated at the time of deployment by
various combinations of length, claw growth ridges (McLaren, 1958), and pelage
characteristics. Seals determined to be less than one year were classified as
young-of-the-year. For those bio-loggers deployed on young-of-the-year and
transmitting into the next year, the age class was advanced to subadult on 1
March of the following year -- the assumed anniversary of their birth. Subadults
are those seals likely greater than one year of age but less than four years.
Adults are individuals that are likely older than four years. Table
\ref{tab:tblDeploy} provides a summary of these deployments and data received
from them.

@@

\setlength{\tabcolsep}{0pt}

@@

\multicolumn{1}{>{\raggedright}m{\dimexpr 0.9in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Spotted\ seal}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.35in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{23\ (21,588\ seal\ hours)}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.35in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{21\ (19,559\ seal\ hours)}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.35in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{11\ (13,417\ seal\ hours)}}} \\

@@

\begin{figure}
\includegraphics[width=1\linewidth]{../figures/examplePlot-1} \caption{\textbf{Example percent-dry actogram from bio-logger data}
\linebreak Haul-out behavior observations recorded by a bio-logger deployed on a
ribbon seal over two years during the months of March, April, and May. This
actogram plot represents the transmitted hourly percent-dry values. Not all data
during a deployment are reliably transmitted from the bio-logger and because data
are transmitted in 24-hour chunks occasional missing days are present in the
record. Here, missing data that were not successfully received from the device
are represented as empty rectangles.}\label{fig:examplePlot}
\end{figure}

@@

\hypertarget{explanatory-variables}{%
\subsection*{Explanatory variables}\label{explanatory-variables}}
\addcontentsline{toc}{subsection}{Explanatory variables}

@@

\hypertarget{haul-out-modeling}{%
\subsection*{Haul-out modeling}\label{haul-out-modeling}}
\addcontentsline{toc}{subsection}{Haul-out modeling}

@@

To properly account for temporal dependence and to take advantage of
computational efficiency, we used generalized linear mixed pseudo-models
(GLMPMs; Ver Hoef, London \& Boveng (2010)) to model variation in
haul-out behavior as a function of (1) covariate predictors, (2) temporally
autocorrelated random effects, and (3) individual random effects representing
heterogeneity in individual behavior. We used the glmmLDTS package
(Ver Hoef, London \& Boveng, 2010) to implement GLMPMs. We explored two different model
formulations for our data and we fit separate models to bearded, ribbon, and
spotted seal data sets as we expected differing behavior by species. Separate
models for each species were also needed because a single, very large data set
proved computationally intractable. In our first model formulation and for each
species, we fitted a year-independent model that predicted average haul-out
behavior as a function of demographic, weather, seasonal, and diurnal effects.
Second, for ribbon and spotted seals (which had considerably more data than
bearded seals), we fitted models that included all the effects from the first
model, but also permitted annual variation in haul-out timing. This second set
of models was used to examine whether haul-out patterns varied by year and to
determine the annual timing of apparent peaks in haul-out behavior. For both
models, we assumed an hourly Bernoulli response (a binary 0-1 response dependent
upon whether the tag was greater than 50\% dry for a given hour) where the linear
predictor was modeled on the logit scale. This is
consistent with previous approaches (London et al., 2012; Ver Hoef et al., 2014) and only
6.995\% of our
observations fell between 10\% and 90\% hourly percent-dry.

@@

\hypertarget{results}{%
\section*{Results}\label{results}}
\addcontentsline{toc}{section}{Results}

@@

\begin{figure}
\includegraphics[width=1\linewidth]{../figures/dataMap-1} \caption{\textbf{Spatial distribution of haul-out records from 1 March through July 15 for bearded, ribbon, and spotted seal.} \linebreak Linking location estimates with haul-out records in space and time allows for inclusion of weather covariates in the final model. For this visualization, data were collated across all years between 2005 and 2020 and each hexagonal cell represents an area of 50 km\textsuperscript{2}}\label{fig:dataMap}
\end{figure}

\begin{figure}
\includegraphics[width=1\linewidth]{../figures/dataCal-1} \caption{\textbf{Seasonal distribution of haul-out behavior observations by species} \linebreak Distribution of hourly percent-dry bio-logger data from 1 March to 15 July for each species. Data are grouped by day-of-year and presented in seal-hours collated across all years between 2005 and 2020. Seal-hours of data are represented as a color gradient and bar height. The higher amount of data from May and June in ribbon and spotted seals coincides with peak molting when seals (and their attached bio-loggers) are more likely hauled out. Additionally, many bio-logger deployments started in April and May. The overall reduced quantity of observations from bearded seals is reflective of the lower number of bio-logger deployments in the study and the fact that the majority of deployments on bearded seals occurred in August and September; therefore most bio-loggers deployed had exhausted their batteries.}\label{fig:dataCal}
\end{figure}

@@

\hypertarget{bearded-seals}{%
\subsection*{Bearded Seals}\label{bearded-seals}}
\addcontentsline{toc}{subsection}{Bearded Seals}

Age and sex class were not included in the model for bearded seals due to our
lower sample size for adult and young-of-year age classes. As such, results are
shown for all ages (Figure \ref{fig:beardedHOCal}; see also
\ref{fig:beardedPredSE}). Additionally, after approximately 9 months, 7 devices
deployed on the rear flipper of bearded seals reported implausible hourly
percent-dry data (100\% dry for several weeks but indicative of movement and
increasing transmission rates (see Boveng \& Cameron (2013))). All data after the first
instance of unrealistic values were censored from this analysis. In addition to
a peak around local solar noon, the bearded seal model predicted additional
haul-out activity around local midnight. In concert with the lower magnitude of
haul-out probability, bearded seal haul-out behavior was also more protracted
throughout the spring season compared to ribbon (Figure \ref{fig:ribbonHOCal})
and spotted seals (Figure \ref{fig:spottedHOCal}). Overall, bearded seals were
less likely to haul out and had a bi-modal distribution of haul-out probability
across the day.

\begin{figure}
\includegraphics[width=1\linewidth]{../figures/beardedHOCal-1} \caption{\textbf{Bearded seal predicted haul-out probability} \linebreak Predicted hourly haul-out probability of bearded seals (all ages and sex classes) from 15 March to 30 June for all age and sex classes combined. Predictions in July and before 15 March were not included to avoid spurious model predictions at the edge of the data range. Weather covariate values in the prediction were based on a simple generalized additive model for each weather covariate with smooth terms for day-of-year and solar hour to account for anticipated variability within a day over the season. Note, our model predicts lower haul-out probability for bearded seals overall compared to ribbon and spotted seals and the color scales are not directly comparable.}\label{fig:beardedHOCal}
\end{figure}

@@

\begin{figure}
\includegraphics[width=1\linewidth]{../figures/beardedHOwx-1} \caption{\textbf{Influence of weather covariates on bearded seal haul-out probability} \linebreak Marginal effects of temperature, wind, barometric pressure, and precipitation on the predicted haul-out probability of bearded seals combined across all age and sex classifications. Hour of the day was fixed at local solar noon and day-of-year fixed at 15 May. The other weather covariate values were predicted from a simple generalized additive model for each weather covariate with smooth terms for day-of-year and solar hour to account for anticipated variability within a day over the season. While the marginal effect of the covariate is continuous, points and vertical lines representing the 95\% confidence interval around the predicted haul-out probability are shown only at observed weather values to give an indication of how the observations are distributed across the range of weather values.}\label{fig:beardedHOwx}
\end{figure}

\hypertarget{ribbon-seals}{%
\subsection*{Ribbon Seals}\label{ribbon-seals}}
\addcontentsline{toc}{subsection}{Ribbon Seals}

@@

\begin{figure}
\includegraphics[width=1\linewidth]{../figures/ribbonHOCal-1} \caption{\textbf{Ribbon seal predicted haul-out probability} \linebreak Predicted hourly haul-out probability of ribbon seals from 15 March to 30 June for each age and sex class used in the model. Weather covariate values in the prediction were based on a simple generalized additive model for each weather covariate with smooth terms for day-of-year and solar hour to account for anticipated variability within a day over the season. Predictions for young of the year show their transition from newly weaned pups resting on the ice to more in-water activities. Predictions in July and before 15 March were not included to avoid spurious model predictions at the edge of the data range.}\label{fig:ribbonHOCal}
\end{figure}

@@

\begin{figure}
\includegraphics[width=1\linewidth]{../figures/ribbonHOwx-1} \caption{\textbf{Influence of weather covariates on ribbon seal haul-out probability} \linebreak Marginal effects of temperature, wind, barometric pressure, and precipitation on the predicted haul-out probability of ribbon seals within each age and sex classifications. Hour of the day was fixed at local solar noon and day-of-year fixed at 15 May. The other weather covariate values were predicted from a simple generalized additive model for each weather covariate with smooth terms for day-of-year and solar hour to account for anticipated variability within a day over the season. While the marginal effect of the covariate is continuous, points and vertical lines representing the 95\% confidence interval around the predicted haul-out probability are shown only at observed weather values to give an indication of how the observations are distributed across the range of weather values.}\label{fig:ribbonHOwx}
\end{figure}

\hypertarget{spotted-seals}{%
\subsection*{Spotted Seals}\label{spotted-seals}}
\addcontentsline{toc}{subsection}{Spotted Seals}

@@

\begin{figure}
\includegraphics[width=1\linewidth]{../figures/spottedHOCal-1} \caption{\textbf{Spotted seal predicted haul-out probability} \linebreak Predicted hourly haul-out probability of spotted seals from 1 March to 30 June for each age and sex class used in the model. Weather covariate values in the prediction were based on a simple generalized additive model for each weather covariate with smooth terms for day-of-year and solar hour to account for anticipated variability within a day over the season. As with ribbon seals, predictions for young of the year show their transition from newly weaned pups resting on the ice to more in-water activities. Predictions in July and before 15 March were not included to avoid spurious model predictions at the edge of the data range.}\label{fig:spottedHOCal}
\end{figure}

Spotted seal haul-out behavior was less affected by the weather covariates
compared to ribbon and bearded seals but their influence on the model was still
significant in some cases.
Temperature
(\(F_{1,114807}\)
= 5.462; \(p\) = 0.019), wind
(\(F_{1,114807}\)
= 46.954; \(p\) = \textless0.001),
and barometric pressure
(\(F_{1,114807}\)
= 10.214; \(p\) =
0.001) were all significant. Spotted seals were more likely to be on the ice when
temperatures were relatively warm and less likely to
haul out at higher wind speeds. Wind chill
(\(F_{1,114807}\)
= 0.559; \(p\) =
0.455) and
precipitation
(\(F_{1,114807}\)
= 0.763; \(p\) =
0.382)
were not as influential as the other covariates.
Differences in the magnitude of response between the age-sex classes were
present and consistent across each of the weather covariates (Figure
\ref{fig:spottedHOwx}). There was a consistent ranking of adult males being the most
likely to haul out, followed by adult females, and, then, subadults. This differs
from ribbon seals which showed more overlap between adult males and adult females
and that subadults were most likely to haul out across the presented range of weather
covariates.

\begin{figure}
\includegraphics[width=1\linewidth]{../figures/spottedHOwx-1} \caption{\textbf{Influence of weather covariates on spotted seal haul-out probability} \linebreak Marginal effects of temperature, wind, barometric pressure, and precipitation on the predicted haul-out probability of spotted seals within each age and sex classification. Hour of the day was fixed at local solar noon and day-of-year fixed at 15 May. The other weather covariate values were predicted from a simple generalized additive model for each weather covariate with smooth terms for day-of-year and solar hour to account for anticipated variability within a day over the season. While the marginal effect of the covariate is continuous, points and vertical lines representing the 95\% confidence interval around the predicted haul-out probability are shown only at observed weather values to give an indication of how the observations are distributed across the range of weather values.}\label{fig:spottedHOwx}
\end{figure}

\hypertarget{annual-variation-in-haul-out-timing}{%
\subsection*{Annual variation in haul-out timing}\label{annual-variation-in-haul-out-timing}}
\addcontentsline{toc}{subsection}{Annual variation in haul-out timing}

>>>>

% Pandoc citation processing
%From Pandoc 3.1.8
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

@@

\section*{Introduction}\label{introduction}
\addcontentsline{toc}{section}{Introduction}

@@

\section*{Methods}\label{methods}
\addcontentsline{toc}{section}{Methods}

\subsection*{Data collection}\label{data-collection}
\addcontentsline{toc}{subsection}{Data collection}

For this study we used haul-out behavior data and location estimates from
bio-loggers deployed on bearded, ribbon, and spotted seals in the Bering,
Chukchi, and western Beaufort seas by multiple organizations as part of
collaborative investigations from 2005 through 2021. Seals were captured using
nets and bio-loggers were attached during studies based in coastal communities
or on research ships. Ship-based capture events occurred during spring near the
southern ice edge in the Bering Sea between 2005 and 2018. Land-based capture
events occurred between 2005 and 2020 from May to October, generally between the
Alaska coastal communities of Scammon Bay in the Bering Sea, Utqiaġvik in the
Chukchi Sea, and Nuiqsut in the Beaufort Sea (Supplemental Material, \hyperref[s1]{S1}).
Data from additional deployments along the Kamchatka peninsula in the western
Bering Sea are also included. We refer readers to Figure \ref{fig:deployplot}
and the primary literature for detailed capture and bio-logger attachment
methods (see also Supplemental Material, \hyperref[s1]{S1}). NOAA-led research was
conducted under the authority of Marine Mammal Protection Act Research Permits
Nos. 782-1676, 782-1765, 15126, and 19309 issued by the National Marine
Fisheries Service, and Letters of Assurance of Compliance with Animal Welfare
Act regulations, Nos. A/NW 2010-3 and A/NW 2016-1 from the Alaska Fisheries
Science Center/Northwest Fisheries Science Center Institutional Animal Care and
Use Committee (IACUC). ADF\&G and NSB field work was covered by Research Permits
Nos. 358-1585, 358-1787, 15324, and 20466 and by ADF\&G IACUC permits Nos. 06-16,
09-21, 2014-03, 2015-25, 2016-23, 0027-2017-27, 0027-2018-29, 0027-2019-041.

\begin{figure}
\includegraphics[width=1\linewidth]{../figures/Figure-001} \caption{\textbf{Initial bio-logger deployment areas for bearded, ribbon, and spotted seals in the study between 2005 and 2020 across multiple regions within the Bering, Chukchi, and western Beaufort seas.} \linebreak Solid regions shown for each species are minimum concave polygons buffered by 60 km for enhanced visibility. The larger, shaded region indicates the spatial extent of aerial survey effort to date in the Bering, Chukchi, and Beaufort seas. Deployments were initiated across a range of months but only haul-out data from 1 March to 15 July were included in the analysis. See Figure \ref{fig:dataMap} for the spatial distribution of observed data used in the study and Supplemental Materials \ref{tab:deploymentTable} for additional deployment details.}\label{fig:deployplot}
\end{figure}

We created a subset of haul-out behavior data from
250 bio-loggers deployed on
35 bearded,
110 ribbon, and
105 spotted seals
to include only records from 1 March to 15 July between
2005 and
2022. Bio-loggers were of the
`SPLASH' or `SPOT' family of tags developed by Wildlife Computers (Redmond,
Washington, USA). Deployments consisted of either a single `SPLASH' device, a
single `SPOT' device, or both types. Devices were either adhered to the hair on
the seal or attached through the rear flipper inter-digital webbing. The use of
bio-loggers adhered to the back or head provides some benefits over flipper
mounted devices (e.g.~increased satellite transmittal rates, locations at sea)
but these fall off during the following annual molt, which, depending on
deployment date, limits the duration of haul-out data they provide especially
during the focus months of our study. Additionally, bio-loggers attached to the
head or dorsal region are often dry while the seal is floating at the surface,
inducing a slight positive bias in the hourly percent-dry values reported by the
bio-logger. For this study, in cases where both bio-logger types were deployed,
we preferred hourly percent-dry observations from the flipper tag. All data were
transmitted by the deployed instruments via the Argos satellite network and
location data were either derived from Argos transmissions or transmitted
FastLoc GPS data.

Sex and age class (non-dependent \emph{young-of-the-year}, sexually immature
\emph{subadults}, and mature \emph{adults}) were estimated at the time of deployment by
various combinations of length, claw growth ridges (McLaren, 1958), and pelage
characteristics. Seals determined to be less than one year were classified as
young-of-the-year. For those bio-loggers deployed on young-of-the-year and
transmitting into the next year (6 ribbon seals; 3 spotted seals), the age class
was advanced to subadult on 1 March of the following year -- the assumed
anniversary of their birth. Subadults are those seals likely greater than one
year of age but less than four years. Adults are individuals that are likely
older than four years. Table \ref{tab:tblDeploy} provides a summary of these
deployments and data received from them.

@@

\setlength{\tabcolsep}{2pt}

@@

\multicolumn{1}{>{\raggedright}m{\dimexpr 0.9in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{Spotted\ seal}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{F}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.35in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{23\ (21,970\ seal\ hours)}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.35in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{21\ (19,559\ seal\ hours)}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 1.35in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{9}{9}\selectfont{11\ (13,417\ seal\ hours)}}} \\

@@

\begin{figure}
\includegraphics[width=1\linewidth]{../figures/Figure-002} \caption{\textbf{Example percent-dry actogram from bio-logger data}
\linebreak Haul-out behavior observations recorded by a bio-logger deployed on a
ribbon seal over two years during the months of March, April, and May. This
actogram plot represents the transmitted hourly percent-dry values. Not all data
during a deployment are reliably transmitted from the bio-logger and because data
are transmitted in 24-hour chunks occasional missing days are present in the
record. Here, missing data that were not successfully received from the device
are represented as empty rectangles.}\label{fig:examplePlot}
\end{figure}

@@

\subsection*{Explanatory variables}\label{explanatory-variables}
\addcontentsline{toc}{subsection}{Explanatory variables}

@@

\subsection*{Haul-out modeling}\label{haul-out-modeling}
\addcontentsline{toc}{subsection}{Haul-out modeling}

@@

To properly account for temporal dependence and to take advantage of
computational efficiency, we used generalized linear mixed pseudo-models
(GLMPMs; Ver Hoef, London \& Boveng (2010)) to model variation in
haul-out behavior as a function of (1) covariate predictors, (2) temporally
autocorrelated random effects, and (3) individual random effects representing
heterogeneity in individual behavior. We used the glmmLDTS package
(Ver Hoef, London \& Boveng, 2010) to implement GLMPMs. We explored two different model
formulations for our data and we fit separate models to bearded, ribbon, and
spotted seal data sets as we expected differing behavior by species. Separate
models for each species were also needed because a single, very large data set
proved computationally intractable. In our first model formulation and for each
species, we fitted a year-independent model that predicted average haul-out
behavior as a function of demographic, weather, seasonal, and diurnal effects.
Second, for ribbon and spotted seals (which had considerably more data than
bearded seals), we fitted models that included all the effects from the first
model, but also permitted annual variation in haul-out timing. This second set
of models was used to examine whether haul-out patterns varied by year and to
determine the annual timing of apparent peaks in haul-out behavior. For both
models, we assumed an hourly Bernoulli response (a binary 0-1 response dependent
upon whether the tag was greater than 50\% dry for a given hour) where the linear
predictor was modeled on the logit scale. This is
consistent with previous approaches (London et al., 2012; Ver Hoef et al., 2014) and only
7.005\% of our
observations fell between 10\% and 90\% hourly percent-dry.

@@

\section*{Results}\label{results}
\addcontentsline{toc}{section}{Results}

@@

\begin{figure}
\includegraphics[width=1\linewidth]{../figures/Figure-003} \caption{\textbf{Spatial distribution of haul-out records from 1 March through July 15 for bearded, ribbon, and spotted seal.} \linebreak Linking location estimates with haul-out records in space and time allows for inclusion of weather covariates in the final model. For this visualization, data were collated across all years between 2005 and 2020 and each hexagonal cell represents an area of 50 km\textsuperscript{2}}\label{fig:dataMap}
\end{figure}

\begin{figure}
\includegraphics[width=1\linewidth]{../figures/Figure-004} \caption{\textbf{Seasonal distribution of haul-out behavior observations by species} \linebreak Distribution of hourly percent-dry bio-logger data from 1 March to 15 July for each species. Data are grouped by day-of-year and presented in seal-hours collated across all years between 2005 and 2020. Seal-hours of data are represented as a color gradient and bar height. The higher amount of data from May and June in ribbon and spotted seals coincides with peak molting when seals (and their attached bio-loggers) are more likely hauled out. Additionally, many bio-logger deployments started in April and May. The overall reduced quantity of observations from bearded seals is reflective of the lower number of bio-logger deployments in the study and the fact that the majority of deployments on bearded seals occurred in August and September; therefore most bio-loggers deployed had exhausted their batteries.}\label{fig:dataCal}
\end{figure}

@@

\subsection*{Bearded Seals}\label{bearded-seals}
\addcontentsline{toc}{subsection}{Bearded Seals}

Age and sex class were not included in the model for bearded seals due to our
lower sample size for adult and young-of-year age classes. As such, results are
shown for all ages (Figure \ref{fig:beardedHOCal}; see also
\ref{fig:beardedPredSE}). Additionally, after approximately 9 months, 7 devices
deployed on the rear flipper of bearded seals reported implausible hourly
percent-dry data (100\% dry for several weeks but indicative of movement and
increasing transmission rates (see Boveng \& Cameron (2013))). All data after the first
instance of unrealistic values were censored from this analysis. In addition to
a peak around local solar noon, the bearded seal model predicted additional
haul-out activity around local midnight. In concert with the lower magnitude of
haul-out probability, bearded seal haul-out behavior was also more protracted
throughout the spring season compared to ribbon
and spotted seals (see below). Overall, bearded seals were
less likely to haul out and had a bi-modal distribution of haul-out probability
across the day.

\begin{figure}
\includegraphics[width=1\linewidth]{../figures/Figure-005} \caption{\textbf{Bearded seal predicted haul-out probability} \linebreak Predicted hourly haul-out probability of bearded seals (all ages and sex classes) from 15 March to 30 June for all age and sex classes combined. Predictions in July and before 15 March were not included to avoid spurious model predictions at the edge of the data range. Weather covariate values in the prediction were based on a simple generalized additive model for each weather covariate with smooth terms for day-of-year and solar hour to account for anticipated variability within a day over the season. Note, our model predicts lower haul-out probability for bearded seals overall compared to ribbon and spotted seals and the color scales are not directly comparable.}\label{fig:beardedHOCal}
\end{figure}

@@

\begin{figure}
\includegraphics[width=1\linewidth]{../figures/Figure-006} \caption{\textbf{Influence of weather covariates on bearded seal haul-out probability} \linebreak Marginal effects of temperature, wind, barometric pressure, and precipitation on the predicted haul-out probability of bearded seals combined across all age and sex classifications. Hour of the day was fixed at local solar noon and day-of-year fixed at 15 May. The other weather covariate values were predicted from a simple generalized additive model for each weather covariate with smooth terms for day-of-year and solar hour to account for anticipated variability within a day over the season. While the marginal effect of the covariate is continuous, points and vertical lines representing the 95\% confidence interval around the predicted haul-out probability are shown only at observed weather values to give an indication of how the observations are distributed across the range of weather values.}\label{fig:beardedHOwx}
\end{figure}

\subsection*{Ribbon Seals}\label{ribbon-seals}
\addcontentsline{toc}{subsection}{Ribbon Seals}

@@

\begin{figure}
\includegraphics[width=1\linewidth]{../figures/Figure-007} \caption{\textbf{Ribbon seal predicted haul-out probability} \linebreak Predicted hourly haul-out probability of ribbon seals from 15 March to 30 June for each age and sex class used in the model. Weather covariate values in the prediction were based on a simple generalized additive model for each weather covariate with smooth terms for day-of-year and solar hour to account for anticipated variability within a day over the season. Predictions for young of the year show their transition from newly weaned pups resting on the ice to more in-water activities. Predictions in July and before 15 March were not included to avoid spurious model predictions at the edge of the data range.}\label{fig:ribbonHOCal}
\end{figure}

@@

\begin{figure}
\includegraphics[width=1\linewidth]{../figures/Figure-008} \caption{\textbf{Influence of weather covariates on ribbon seal haul-out probability} \linebreak Marginal effects of temperature, wind, barometric pressure, and precipitation on the predicted haul-out probability of ribbon seals within each age and sex classifications. Hour of the day was fixed at local solar noon and day-of-year fixed at 15 May. The other weather covariate values were predicted from a simple generalized additive model for each weather covariate with smooth terms for day-of-year and solar hour to account for anticipated variability within a day over the season. While the marginal effect of the covariate is continuous, points and vertical lines representing the 95\% confidence interval around the predicted haul-out probability are shown only at observed weather values to give an indication of how the observations are distributed across the range of weather values.}\label{fig:ribbonHOwx}
\end{figure}

\subsection*{Spotted Seals}\label{spotted-seals}
\addcontentsline{toc}{subsection}{Spotted Seals}

@@

\begin{figure}
\includegraphics[width=1\linewidth]{../figures/Figure-009} \caption{\textbf{Spotted seal predicted haul-out probability} \linebreak Predicted hourly haul-out probability of spotted seals from 1 March to 30 June for each age and sex class used in the model. Weather covariate values in the prediction were based on a simple generalized additive model for each weather covariate with smooth terms for day-of-year and solar hour to account for anticipated variability within a day over the season. As with ribbon seals, predictions for young of the year show their transition from newly weaned pups resting on the ice to more in-water activities. Predictions in July and before 15 March were not included to avoid spurious model predictions at the edge of the data range.}\label{fig:spottedHOCal}
\end{figure}

Spotted seal haul-out behavior was less affected by the weather covariates
compared to ribbon and bearded seals but their influence on the model was still
significant in some cases.
Temperature
(\(F_{1,115189}\)
= 5.384; \(p\) = 0.020), wind
(\(F_{1,115189}\)
= 45.718; \(p\) = \textless0.001),
and barometric pressure
(\(F_{1,115189}\)
= 9.445; \(p\) =
0.002) were all significant. Spotted seals were more likely to be on the ice when
temperatures were relatively warm and less likely to
haul out at higher wind speeds. Wind chill
(\(F_{1,115189}\)
= 0.72; \(p\) =
0.396) and
precipitation
(\(F_{1,115189}\)
= 0.773; \(p\) =
0.379)
were not as influential as the other covariates.
Differences in the magnitude of response between the age-sex classes were
present and consistent across each of the weather covariates (Figure
\ref{fig:spottedHOwx}). There was a consistent ranking of adult males being the most
likely to haul out, followed by adult females, and, then, subadults. This differs
from ribbon seals which showed more overlap between adult males and adult females
and that subadults were most likely to haul out across the presented range of weather
covariates.

\begin{figure}
\includegraphics[width=1\linewidth]{../figures/Figure-010} \caption{\textbf{Influence of weather covariates on spotted seal haul-out probability} \linebreak Marginal effects of temperature, wind, barometric pressure, and precipitation on the predicted haul-out probability of spotted seals within each age and sex classification. Hour of the day was fixed at local solar noon and day-of-year fixed at 15 May. The other weather covariate values were predicted from a simple generalized additive model for each weather covariate with smooth terms for day-of-year and solar hour to account for anticipated variability within a day over the season. While the marginal effect of the covariate is continuous, points and vertical lines representing the 95\% confidence interval around the predicted haul-out probability are shown only at observed weather values to give an indication of how the observations are distributed across the range of weather values.}\label{fig:spottedHOwx}
\end{figure}

\subsection*{Annual variation in haul-out timing}\label{annual-variation-in-haul-out-timing}
\addcontentsline{toc}{subsection}{Annual variation in haul-out timing}
